---
title: 'Analyzing OpenAIRE identifiers'
output: 
  html_document: 
    df_print: kable
date: "2022-11-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```

```{r read}
# create table id_prefixes as select type, substring(id, 1, 2) as prefix, substring(id, 4, 12) as schema, count(*) from entities group by 1, 2, 3;
# insert into id_prefixes select reltype || '_source', substring(sourceid, 1, 2) as prefix, substring(sourceid, 4, 12) as schema, count(*) from relations group by 1, 2, 3;
# insert into id_prefixes select reltype || '_target', substring(targetid, 1, 2) as prefix, substring(targetid, 4, 12) as schema, count(*) from relations group by 1, 2, 3;
# select * from id_prefixes;
d = readr::read_csv('ids.csv')
```

# Introduction

Each OpenAIRE id used in the JSON dump consists of three parts:

* a two-digits numeric prefix
* 14-characters long, right padded with underscores name looking like an id namespace or something like that
* 32-characters string looking like a hash

In this document it was explored if there are any systematic relations between those parts which we can exploit to better organize the data inside the database.

# Analyzis

```{r}
d %>% filter(!grepl('_source|_target', type)) %>% group_by(type, prefix) %>% summarize(n = n()) %>% tidyr::pivot_wider('type', names_from = 'prefix', values_from = 'n')
```

The id prefix seems to determine the entity schema:

* `00` - community_infrastracture
* `10` - datasource
* `20` - organization
* `40` - project
* `50` - research product (dataset/publication/software/other research product)

```{r}
d = d %>%
  mutate(typeId = case_when(
    prefix == 0 ~ 'communities_infrastructures',
    prefix == 10 ~ 'datasource',
    prefix == 20 ~ 'organization',
    prefix == 40 ~ 'project',
    prefix == 50 ~ 'research product'
  ))
```

Knowing that we can determine domain and range of relation properties.

First let's take a look at domains (numbers are relations count).

```{r}
d %>% filter(grepl('_source', type)) %>% group_by(type, typeId) %>% summarize(n = sum(count)) %>% tidyr::pivot_wider('type', names_from = 'typeId', values_from = 'n') %>% arrange(type)
```

Doesn't look bad. Most properties have single domain. The only exceptions are:

* `isProvidedBy` having domain `research product` or `datasource`
* `provides` having domain `datasource` or `ogranization`
* `IsRelatedTo` having any entity type as a domain

Same table for ranges:

```{r}
d %>% filter(grepl('_target', type)) %>% group_by(type, typeId) %>% summarize(n = sum(count)) %>% tidyr::pivot_wider('type', names_from = 'typeId', values_from = 'n') %>% arrange(type)
```

With predictable outcomes:

* `isProvidedBy` having range `datasource` or `ogranization`
* `provides` having range `research product` or `datasource`
* `IsRelatedTo` having any entity type as a range
* all other properties having single entity type as a range.

Now let's find out how many of the id prefixes (combination of entity type and namespace part) appearing in the relations actually exist for entities.

Domain side:

```{r}
de = d %>% filter(!grepl('_source|_target', type)) %>% mutate(exists = TRUE) %>% select(-count)
da = d %>% 
  filter(grepl('_source', type)) %>%
  rename(relation = type, relCount = count) %>%
  mutate(relation = sub('_.*', '', relation)) %>%
  left_join(de) %>%
  group_by(relation, prefix, typeId, exists) %>%
  summarize(relCount = sum(relCount)) %>%
  arrange(relation, exists, typeId) %>%
  rename(domain = 'typeId')
da
da %>% group_by(exists) %>% summarize(relCount = sum(relCount)) %>% ungroup() %>% mutate(percentage = round(100 * relCount / sum(relCount), 1))
```

Range side:

```{r}
de = d %>% filter(!grepl('_source|_target', type)) %>% mutate(exists = TRUE) %>% select(-count)
ra = d %>% 
  filter(grepl('_target', type)) %>%
  rename(relation = type, relCount = count) %>%
  mutate(relation = sub('_.*', '', relation)) %>%
  left_join(de) %>%
  group_by(relation, prefix, typeId, exists) %>%
  summarize(relCount = sum(relCount)) %>%
  arrange(relation, exists, typeId) %>%
  rename(range = 'typeId')
ra
ra %>% group_by(exists) %>% summarize(relCount = sum(relCount)) %>% ungroup() %>% mutate(percentage = round(100 * relCount / sum(relCount), 1))
```

While we can see some id namespaces used in relations are missing in entities, the scale is neglectably small (0.4% of relations count) and skipping them on import won't provide any measurable speedup.

Finally let's check if the hash part of the id is unique and can be used alone.

```sql
> select substring(id, 18), count(*) from entities group by 1 having count(*) > 1;
(...)
(531496 rows)
```

No, hashes alone aren't unique.